---
title: "🔥 Wildfire Cartography: Visualizing NASA FIRMS DataR"
author: "DS 421 – Cartographic Design"
date: "2025"
format:
  html:
    toc: true
    code-fold: true
    theme: minty
    css: styles.css
    self-contained: false
execute:
  eval: false
---

------------------------------------------------------------------------

Welcome to your hands-on lab using **NASA FIRMS wildfire detection data**. In this tutorial, you will:

-   Explore spatial file formats: shapefiles (.shp) and KMZ/KML
-   Visualize fire detections using `ggplot2`, `mapboxgl`, `leaflet`, and `gganimate`
-   Learn about satellite-based thermal sensing and what the data columns mean
-   Practice both static and interactive map design

You’ll work with **real satellite data** capturing active fires detected across the continental U.S. and Hawaii by LANDSAT.

------------------------------------------------------------------------

## 📆 Load Required Libraries

```{r}
library(tidyverse)
library(sf)            # Read and handle vector spatial data
library(here)          # File paths
library(dplyr)         # Data wrangling
library(ggplot2)       # Static visualizations
library(mapgl)      # Interactive mapping
library(rnaturalearth) # Country/State boundaries
library(gganimate)     # Animations over time
library(leaflet)       # Lightweight web maps
```

------------------------------------------------------------------------

## 📐 About the Data

The data comes from [NASA FIRMS](https://firms.modaps.eosdis.nasa.gov/usfs/active_fire/#firms-shapefile), which provides **thermal anomaly** and **fire detection data** derived from satellite observations. Each point represents a heat signature flagged as a likely fire, with confidence scores:

-   `low`: Possibly noise or cloud
-   `medium`: Reasonable detection
-   `high`: Very likely a fire

In this tutorial, we use the **7-day LANDSAT active fire shapefile** and its corresponding **KMZ fire footprint file**.

------------------------------------------------------------------------

## 📍 Read Point-Based Fire Detection Shapefile

On this website: <https://firms.modaps.eosdis.nasa.gov/usfs/active_fire/#firms-shapefile:~:text=Real%2DTime%20data.-,Shapefiles,-Active%20fire%20detection> under Shapefiles, download the LANDSAT 30M USA (Conterminous) and Hawaii 7 day zip file

-   Click on the zip file you downloaded to unzip it

-   Create a data folder within DS422-Geocomputation and move the

-   Move the unzipped file (will be a folder with 5 files within it) into the new data folder you created

-   Now you will be able to read the file in with the code below

```{r}
wildfire_shp_data <- st_read(here("data/LANDSAT_USA_contiguous_and_Hawaii_7d/LANDSAT_USA_contiguous_and_Hawaii_7d.shp"))
```

This dataset contains point geometries where fires were detected. Important columns include:

-   `ACQ_DATE`: Date of detection
-   `LATITUDE`, `LONGITUDE`: Coordinates
-   `CONFIDENCE`: Detection confidence category (low, medium, high)

------------------------------------------------------------------------
```{r}
DS422-Geocomputation/
  data/
  wildfire_firms.qmd   (this file)
  styles.css           (optional)

```

## 📈 Challenge 1

Use ggplot to create a line plot of fires with the day on the x axis and number of fires on the y axis

```{r}
library(lubridate)

fires_by_day <- wildfire_shp_data |>
  mutate(day = as.Date(ACQ_DATE)) |>
  count(day, name = "n") |>
  arrange(day)

ggplot(fires_by_day, aes(x = day, y = n)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  labs(
    title = "Fires detected per day (LANDSAT, last 7 days)",
    x = "Date", y = "Number of detections",
    caption = "Source: NASA FIRMS (LANDSAT 30m)"
  ) +
  theme_minimal()

```

```{r}

```

------------------------------------------------------------------------

## 📈 Challenge 2

Use ggplot to create a stacked bar chart colored by category with number of fires on the y axis for each day

```{r}
fires_by_day_conf <- wildfire_shp_data %>%
  st_drop_geometry() %>%
  group_by(ACQ_DATE, CONFIDENCE) %>%
  summarise(num_fires = n(), .groups = "drop") %>%
  arrange(ACQ_DATE)

ggplot(fires_by_day_conf, aes(x = ACQ_DATE, y = num_fires, fill = CONFIDENCE)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Daily Wildfire Detections by Confidence Level",
    x = "Date of Detection",
    y = "Number of Fires",
    fill = "Confidence Level"
  ) +
  theme_minimal()
```

## 📏 Map Fire Points by Confidence Level

```{r}
confidence_colors <- c(
  "low" = "#1a9850",
  "medium" = "#fee08b",
  "high" = "#d73027"
)

mapboxgl(bounds = wildfire_shp_data) |>
  add_circle_layer(
    id = "wildfires_categorical",
    source = wildfire_shp_data,
    circle_color = match_expr(
      column = "CONFIDENCE",
      values = names(confidence_colors),
      stops = unname(confidence_colors)
    ),
    circle_radius = 4,
    circle_opacity = 0.7,
    popup = "ACQ_DATE",
    tooltip = "CONFIDENCE",
    hover_options = list(
      circle_color = "black",
      circle_radius = 6
    )
  ) |>
  add_legend(
    "Confidence Level",
    values = names(confidence_colors),
    colors = unname(confidence_colors),
    type = "categorical"
  )
```

------------------------------------------------------------------------

## 🔥 Create a Heatmap of Fire Activity

```{r}
mapboxgl(style = mapbox_style("dark"), bounds = wildfire_shp_data) |>
  add_heatmap_layer(
    id = "fire_heatmap",
    source = wildfire_shp_data,
    heatmap_color = interpolate(
      property = "heatmap-density",
      values = seq(0, 1, 0.2),
      stops = c(
        "rgba(33,102,172,0)", "rgb(103,169,207)",
        "rgb(209,229,240)", "rgb(253,219,199)",
        "rgb(239,138,98)", "rgb(178,24,43)"
      )
    ),
    heatmap_opacity = 0.4
  )
```

```{r}
mapboxgl(style = mapbox_style("dark"), bounds = wildfire_shp_data) |>
  add_heatmap_layer(
    id = "heatmap",
    source = wildfire_shp_data,
    heatmap_radius = 3, 
    heatmap_color = interpolate(
      property = "heatmap-density",
      values = seq(0, 1, 0.2),
      stops = c("transparent", viridisLite::viridis(5))
    )
  )
```

------------------------------------------------------------------------

## 🎙️ Animate Fire Detection Over Time

```{r}
usa <- ne_states(country = "united states of america", returnclass = "sf")

# Plot animated wildfire data
p <- ggplot() +
  geom_sf(data = usa, fill = "grey95", color = "white") +
  geom_point(
    data = wildfire_shp_data,
    aes(x = LONGITUDE, y = LATITUDE, color = CONFIDENCE, group = ACQ_DATE),
    alpha = 0.8,
    size = 1.2
  ) +
  scale_color_manual(
    values = c("low" = "forestgreen", "medium" = "orange", "high" = "red")
  ) +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50)) +
  theme_minimal() +
  labs(
    title = "🔥 Wildfire Detections Across the U.S.",
    subtitle = "Date: {frame_time}",
    color = "Confidence",
    caption = "Source: LANDSAT Fire Detections"
  ) +
  transition_time(as.Date(wildfire_shp_data$ACQ_DATE)) +
  ease_aes('linear') 

animate(
  p,
  nframes = 30,
  fps = 5,
  width = 800,
  height = 600,
  renderer = gifski_renderer(loop = TRUE)
)

```

------------------------------------------------------------------------

## 🛡️ Load and Visualize KML Fire Area Polygons

-   Download the FirespotArea_LANDSAT_USA_contiguous_and_Hawaii_7d.kmz from here: <https://firms.modaps.eosdis.nasa.gov/usfs/active_fire/#firms-shapefile:~:text=Fire%20detection%20location-,Active%20fire%20detection%20footprint%20and%20centroid%20KMLs/KMZs,-Download%20MODIS%2C%20VIIRS> under **Active fire detection footprint and centroid KMLs/KMZs download the LANDSAT 30m** USA (Conterminous) and Hawaii 7d

-   Move the .kmz file into your data folder

-   The code below will unzip it, move it into a fire_kml file directory that contains a kml file that is readable in R

```{r}
library(sf)
library(leaflet)

# 1) Point to your folder
kml_dir <- "/Users/mckenzietimsing/Desktop/DS422/DS422-Geocomputation/data/fire_kml"

if (!dir.exists(kml_dir)) stop("Folder not found: ", kml_dir)

# 2) Find the KML file that was unzipped from the KMZ
kml_files <- list.files(kml_dir, pattern = "\\.kml$", full.names = TRUE, ignore.case = TRUE)
if (length(kml_files) == 0) stop("No .kml found in: ", kml_dir, 
                                 "\nRe-unzip the KMZ into this folder.")

kml_path <- kml_files[1]
message("Using KML: ", kml_path)

# 3) Discover layer names (KML “folders”)
layer_names <- sf::st_layers(kml_path)$name
print(layer_names)

# Prefer polygon footprints if present
footprint_layer <- layer_names[grepl("Footprint", layer_names, ignore.case = TRUE)][1]
if (is.na(footprint_layer)) footprint_layer <- layer_names[1]

# 4) Read, clean, project
wildfire_area_data <- sf::st_read(kml_path, layer = footprint_layer, quiet = TRUE) |>
  sf::st_zm(drop = TRUE, what = "ZM") |>
  sf::st_make_valid() |>
  sf::st_transform(4326)

# 5) Quick map
leaflet(wildfire_area_data) |>
  addProviderTiles("CartoDB.DarkMatter") |>
  addPolygons(color = "#FF3B30", weight = 0.8, opacity = 0.9)

```

------------------------------------------------------------------------

## 📈 Challenge 3

Download the LANDSAT 30m CSV for USA (Conterminous) and Hawaii at the bottom right of this page: <https://firms.modaps.eosdis.nasa.gov/usfs/active_fire/#firms-shapefile>

-   Move it into the data folder

-   Read the csv into R as a dataframe

-   Transform the dataframe into an sf object using the function [st_as_sf()](https://r-spatial.github.io/sf/reference/st_as_sf.html) so it can be mapped. As another argument within the same [st_as_sf(](https://r-spatial.github.io/sf/reference/st_as_sf.html)) function assign a coordinate reference system.

-   Map the data with any package you would like, just make sure there is at least a slight difference from the examples above, be as creative as possible while utilizing your critical thinking

```{r}
library(readr)
library(dplyr)
library(sf)
library(here)

# Find the FIRMS CSV you saved in data/
csv_path <- list.files(
  here("data"),
  pattern = "LANDSAT.*USA.*Hawaii.*7d.*\\.csv$",
  full.names = TRUE,
  ignore.case = TRUE
)
if (length(csv_path) == 0) stop("Couldn't find the LANDSAT 30m 7-day CSV in data/.")
message("Using CSV: ", csv_path[1])

# Read as a regular dataframe (readr is fast & handles types well)
fires_df <- readr::read_csv(csv_path[1], show_col_types = FALSE)

# Normalize column names to uppercase so we can rely on names
fires_df <- fires_df |> rename_with(toupper)

# Minimal QC: keep rows with valid coordinates
fires_df <- fires_df |> filter(!is.na(LONGITUDE), !is.na(LATITUDE))

# Make sure date is a real Date
if ("ACQ_DATE" %in% names(fires_df)) {
  fires_df <- fires_df |> mutate(ACQ_DATE = as.Date(ACQ_DATE))
}

# To avoid surprises, create a lowercase confidence column if present
if ("CONFIDENCE" %in% names(fires_df)) {
  fires_df <- fires_df |> mutate(CONF_LC = tolower(CONFIDENCE))
} else {
  fires_df <- fires_df |> mutate(CONF_LC = NA_character_)
}

# Convert to sf (keep LONG/LAT columns around for filtering/plotting)
fires_sf <- fires_df |>
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326, remove = FALSE)

fires_sf

library(leaflet)

# Simple palette for confidence (falls back to gray if missing/unknown)
pal <- colorFactor(
  palette = c("low" = "#1a9850", "medium" = "#fee08b", "high" = "#d73027"),
  domain  = c("low","medium","high"),
  na.color = "#9e9e9e"
)

# Split into overlay groups for toggling
fires_low    <- fires_sf |> filter(CONF_LC == "low")
fires_medium <- fires_sf |> filter(CONF_LC == "medium")
fires_high   <- fires_sf |> filter(CONF_LC == "high")
fires_other  <- fires_sf |> filter(is.na(CONF_LC) | !(CONF_LC %in% c("low","medium","high")))

leaflet() |>
  addProviderTiles("Esri.WorldImagery", group = "Imagery") |>
  addProviderTiles("CartoDB.Positron", group = "Light") |>
  addCircleMarkers(data = fires_high,
                   ~LONGITUDE, ~LATITUDE,
                   radius = 4, stroke = FALSE, fillOpacity = 0.8,
                   color = ~pal(CONF_LC),
                   popup = ~paste0("<b>Date:</b> ", ACQ_DATE, 
                                   if ("SATELLITE" %in% names(fires_sf)) paste0("<br><b>Sat:</b> ", SATELLITE) else "",
                                   if ("CONFIDENCE" %in% names(fires_sf)) paste0("<br><b>Conf:</b> ", CONFIDENCE) else ""),
                   clusterOptions = markerClusterOptions(),
                   group = "High") |>
  addCircleMarkers(data = fires_medium,
                   ~LONGITUDE, ~LATITUDE,
                   radius = 4, stroke = FALSE, fillOpacity = 0.8,
                   color = ~pal(CONF_LC),
                   popup = ~paste0("<b>Date:</b> ", ACQ_DATE,
                                   if ("SATELLITE" %in% names(fires_sf)) paste0("<br><b>Sat:</b> ", SATELLITE) else "",
                                   if ("CONFIDENCE" %in% names(fires_sf)) paste0("<br><b>Conf:</b> ", CONFIDENCE) else ""),
                   clusterOptions = markerClusterOptions(),
                   group = "Medium") |>
  addCircleMarkers(data = fires_low,
                   ~LONGITUDE, ~LATITUDE,
                   radius = 4, stroke = FALSE, fillOpacity = 0.8,
                   color = ~pal(CONF_LC),
                   popup = ~paste0("<b>Date:</b> ", ACQ_DATE,
                                   if ("SATELLITE" %in% names(fires_sf)) paste0("<br><b>Sat:</b> ", SATELLITE) else "",
                                   if ("CONFIDENCE" %in% names(fires_sf)) paste0("<br><b>Conf:</b> ", CONFIDENCE) else ""),
                   clusterOptions = markerClusterOptions(),
                   group = "Low") |>
  addCircleMarkers(data = fires_other,
                   ~LONGITUDE, ~LATITUDE,
                   radius = 4, stroke = FALSE, fillOpacity = 0.8,
                   color = ~pal(CONF_LC),
                   popup = ~paste0("<b>Date:</b> ", ACQ_DATE),
                   clusterOptions = markerClusterOptions(),
                   group = "Other/NA") |>
  addLayersControl(
    baseGroups = c("Imagery", "Light"),
    overlayGroups = c("High", "Medium", "Low", "Other/NA"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  addLegend(
    "bottomright", pal = pal, values = c("low","medium","high"),
    title = "Confidence", opacity = 1
  ) |>
  fitBounds(lng1 = min(fires_sf$LONGITUDE, na.rm = TRUE),
            lat1 = min(fires_sf$LATITUDE,  na.rm = TRUE),
            lng2 = max(fires_sf$LONGITUDE, na.rm = TRUE),
            lat2 = max(fires_sf$LATITUDE,  na.rm = TRUE))
coord_sf(xlim = c(-170, -65), ylim = c(18, 72))


```


## ✨ Summary

-   You worked with **shapefiles**, CSV and **KML** for geospatial fire data
-   You mapped **points**, **heatmaps**, and **animated detections**
-   You explored how **satellite sensors** power near real-time cartography

## ✨ Advanced

-   Can you pull Satellite Data for a day in our timeframe and check the predictions for smoke/other fire markers

-   What are the challenges

-   What are some other test we can perform to confirm these fire predictions

```{r}


```

```         
```
